The role of cognateness in non-native spoken word recognition

### 2024-05-09 manuscript/manuscript.tex
Add Methods section to Experiment 3

\begin{longtable}{l|rrrrrrc}

@@

\toprule
\multicolumn{1}{l}{} & \multicolumn{2}{c}{Sample size} & \multicolumn{3}{c}{Age} & \multicolumn{2}{c}{L2} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-6} \cmidrule(lr){7-8}
\multicolumn{1}{l}{} & N & Excluded & M & SD & Range & N knows L2? & Which L2? \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 35 & 8 & $21.80$ & $2.08$ & $18$–$26$ & 1 & Russian \\ 
cat-ENG & 36 & 4 & $21.72$ & $2.25$ & $18$–$25$ & 5 & German, Punjabi, French, Italian, Several \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & 33 & 12 & $21.85$ & $3.00$ & $18$–$33$ & 12 & Italian, French, German \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 32 & 1 & $21.72$ & $2.59$ & $18$–$26$ & 3 & Japanese, German \\ 
cat-ENG & 32 & 0 & $22.31$ & $2.39$ & $18$–$26$ & 2 & Cantonese, Irish \\ 
\bottomrule

@@

@@

We arranged two lists of input words to be presented to participants in
the auditory modality: one in Catalan and one in Spanish. Words in the
Catalan list were 5.02 phonemes long on average (\emph{SD} = 1.49,
\emph{Range} = 2-8), and the orthographic form of their English
translations (which participants had to type), were 5.12 characters long
on average (\emph{SD} = 1.56, \emph{Range} = 3-9. Words in the Spanish
list were 5.52 phonemes long on average (\emph{SD} = 1.47, \emph{Range}
= 3-9), and the orthographic form of their English translations (which
participants had to type), were 5.29 characters long on average
(\emph{SD} = 1.77, \emph{Range} = 3-12.

Participants listened to one audio file in each trial, each containing a
single word presented in isolation. The audio files were the same ones
used in child experiments conducted in the Laboratori de Recerca en
Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio
files were recorded by a proficient Catalan-Spanish female bilingual
from the Metropolitan Area of Barcelona in a child-directed manner.
Catalan and Spanish words were recorded at 44,100 Hz in separate files
in the same session, and then de-noised using Audacity and normalised at
peak intensity using Praat (\citeproc{ref-broersma2021praat}{Broersma \&
Weenink, 2021}). The average duration of the Catalan audio files was
1.24 (\emph{SD} = 0.19, \emph{Range} = 0.8-1.58). The average duration
of the Catalan audio files was 1.16 (\emph{SD} = 0.15, \emph{Range} =
0.78-1.53).

For each translation pair, we defined three predictors of interest: the
lexical frequency of the correct translation (\emph{Frequency}), the
phonological similarity between the presented (Catalan or Spanish) word
and their correct English translation (\emph{Cognateness}), and
presented word's number of cross-language phonological neighbours
(\emph{CLPN}).

We included \emph{Frequency} under the hypothesis that, keeping other
predictors constant, participants would translate higher-frequency words
more accurately and faster than lower-frequency words. Lexical
frequencies of correct translations were extracted from SUBTLEX-UK
(\citeproc{ref-van2014subtlex}{Van Heuven et al., 2014}), and
transformed to Zipf scores (\citeproc{ref-van2014subtlex}{Van Heuven et
al., 2014}). Words in the stimuli list without a lexical frequency value
were excluded from data analysis (2 in the Catalan list, 5 in the
Spanish list).

We calculated \emph{Cognatenesss}, our main predictor of interest, by
computing the Levenshtein similarity between the X-SAMPA transcriptions
of each pair of translations using the \texttt{stringdist} R package
(\citeproc{ref-van2014stringdist}{van der Loo, 2014}). The Levenshtein
distance computes the edit distance between two character strings (in
this case, phoneme transcriptions) by counting the number of additions,
deletions, and substitutions necessary to make both strings identical
(\citeproc{ref-levenshtein1966binary}{Levenshtein et al., 1966}). We
divided this edit distance by the number of characters included in the
longest X-SAMPA transcription of the translation pair. This results in a
proportion score, in which values closer to zero correspond to lower
Levenshtein distances between phonological transcriptions (i.e., higher
similarity), and values closer to 1 correspond to higher Levenshtein
distances (i.e., lower similarity). This transformation also help
account for the fact that the Levenshtein distance tends to increase
with the length of the transcriptions. For interpretability, we
subtracted this proportion from one, so that values closer to one
correspond to higher similarity between phonological transcriptions, and
values closer to zero correspond to lower similarity between
phonological transcriptions. For example, the \emph{table}
(teɪbəl)-\emph{mesa} (mesa) translation pair had a 17\% similarity,
while the \emph{train} (trEIn)-\emph{tren} (tɾEn) translation pair had a
60\% similarity. Figure 1 summarises the lexical frequency, phonological
neighbourhood density and phonological overlap of the words included in
the Catalan and the Spanish lists.

For each Catalan and Spanish word, we calculated the number of
\emph{CLPN} by counting the number of English words with same or higher
lexical frequency, and whose phonological transcription (in X-SAMPA
format) different in up to one phoneme from that of the presented
Catalan or Spanish word. Lexical frequencies and phonological
transcriptions were extracted from the multilingual database CLEARPOND
(\citeproc{ref-marian2012clearpond}{Marian et al., 2012})\footnote{{[}Phonological
  trnascriptions in CLEARPOND were generated from eSPEAK,
  \url{http://espeak.sourceforge.net/}{]}}.

@@

\caption{\label{tbl-stimuli}}

@@

\toprule
\multicolumn{1}{l}{} & \multicolumn{2}{c}{Frequency} & \multicolumn{2}{c}{Cognateness} & \multicolumn{2}{c}{CLPN} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
\multicolumn{1}{l}{} & Mean ± SD & Range & Mean ± SD & Range & Mean ± SD & Range \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $5.88$ ± $0.57$ & $4.43$–$7.24$ & $0.13$ ± $0.18$ & $0.00$–$0.75$ & $0.19$ ± $0.79$ & $0$–$5$ \\ 
cat-ENG & $5.89$ ± $0.59$ & $4.43$–$7.27$ & $0.16$ ± $0.18$ & $0.00$–$0.67$ & $0.76$ ± $2.53$ & $0$–$15$ \\ 
cat-SPA & $5.79$ ± $0.63$ & $4.48$–$7.70$ & $0.38$ ± $0.26$ & $0.00$–$1.00$ & $0.87$ ± $2.08$ & $0$–$12$ \\ 
\bottomrule

@@

Each trial started with a yellow fixation point presented during one
second on the centre of the screen over a black background. After one
second, the audio started playing while the dot remained being displayed
until the audio ended. Upon the offset of the fixation point and audio,
participants were prompted to write their answer by a
\texttt{\textgreater{}} symbol. Typed letters were displayed in the
screen in real time to provide visual feed-back to participants.
Participants were allowed to correct their answer. Then, participants
pressed the \texttt{RETURN/ENTER} key to confirm their answer and start
and new trial. We excluded trials where the participant's answer was not
a valid word in the target language or no answer was given. Trials in
which the response was mistyped by only one character were counted as
correct (e.g., ``pengiun'' instead of ``penguin''), as long as the
response did not correspond to a distinct word. Trials in which
participants took longer than 10 seconds to respond were also excluded.
Participants contributed a total of 9152 valid trials (5694 in Catalan,
3458 in Spanish). The task took approximately 15 minutes to be
completed.

@@

{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{../img/design.png}

@@

\caption{Schematic representation of a trial in the experimental task.}

@@

@@

We collected data for a total of 6,446 trials completed by 72 unique
participants. Of those trials, 2,988 were provided by 36 unique
participants who listened to Catalan words, and 3,458 trials were
provided by 36 unique participants who listened to Spanish words. We
excluded trials in which participants did not enter any text (\emph{n} =
72), in which a response in a language other than English was provided
(e.g., \texttt{agua}, \emph{n} = 51), in which participants did not
provide a whole word (e.g., \texttt{f}, \emph{n} = 5), and in which
participants added comments to the experimenter (e.g., \texttt{unsure},
(\emph{n} = 13). In addition, we excluded data from participants that
self-rated their oral and/or written skills in Catalan and Spanish, or
any other second language as four or higher in a five-point scale
(\emph{n} = 2), were diagnosed with a language (\emph{n} = 1), or did
not contribute more than 80\% of valid trials (\emph{n} = 9). After
applying trial-level and participant-level inclusion criteria, the
resulting dataset included 5,204 trials provided by 54 unique
participants. Of those trials, 2,602 were provided by 27 unique
participants who listened to Catalan words, and 2,604 trials were
provided by 32 unique participants who listened to Spanish words.
Responses given by English participants to Catalan presented words were
5.35 characters long on average (\emph{Median} = 5, \emph{SD} = 1.79,
Range = 1-14), while their translations to Spanish responses were 5.57
characters long on average (\emph{Median} = 5, \emph{SD} = 1.97, Range =
2-21).

@@

\begin{longtable}{l|rrrr}

@@

\toprule
\multicolumn{1}{l}{} & \multicolumn{2}{c}{Correct responses} & \multicolumn{2}{c}{Incorrect responses} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
\multicolumn{1}{l}{} & Correct & Typo & Wrong translation & False friend \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
cat-ENG & $429$ ($16.47\%$) & $11$ ($0.42\%$) & $2,082$ ($79.95\%$) & $82$ ($3.15\%$) \\ 
spa-ENG & $374$ ($14.37\%$) & $2$ ($0.08\%$) & $2,117$ ($81.36\%$) & $109$ ($4.19\%$) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & $780$ ($46.93\%$) & $20$ ($1.20\%$) & $736$ ($44.28\%$) & $126$ ($7.58\%$) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
cat-ENG & $477$ ($18.01\%$) & $7$ ($0.26\%$) & $1,986$ ($75.00\%$) & $178$ ($6.72\%$) \\ 
spa-ENG & $590$ ($19.53\%$) & $6$ ($0.20\%$) & $2,294$ ($75.94\%$) & $131$ ($4.34\%$) \\ 
\bottomrule

@@

\begin{longtable}{l|rrrrrrr}

@@

\toprule
\multicolumn{1}{l}{} &  & \multicolumn{3}{c}{Accuracy (\%)} & \multicolumn{3}{c}{Valid trials} \\ 
\cmidrule(lr){3-5} \cmidrule(lr){6-8}
\multicolumn{1}{l}{} & N & Mean ± SD & SE & Range & Mean ± SD & N trials & Range \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $27$ & $15.86$ ± $5.20$ & $3.05$ & $8.82$–$28.71$ & $96.37$ ± $2.88$ ± $2.88$ & $2,602$ & $87$–$98$ \\ 
cat-ENG & $32$ & $18.48$ ± $4.89$ & $3.27$ & $10.84$–$32.56$ & $81.38$ ± $3.17$ ± $3.17$ & $2,604$ & $71$–$83$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & $21$ & $48.30$ ± $5.29$ & $10.54$ & $38.27$–$58.97$ & $79.14$ ± $3.14$ ± $3.14$ & $1,662$ & $72$–$82$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $31$ & $20.92$ ± $8.29$ & $3.76$ & $5.88$–$44.12$ & $97.45$ ± $1.80$ ± $1.80$ & $3,021$ & $88$–$98$ \\ 
cat-ENG & $32$ & $19.74$ ± $4.94$ & $3.49$ & $10.34$–$27.91$ & $82.75$ ± $0.44$ ± $0.44$ & $2,648$ & $82$–$83$ \\ 
\bottomrule

@@

\captionsetup{labelsep=none}\includegraphics{manuscript_files/figure-pdf/fig-epreds-1-1.pdf}

@@

\caption{\label{fig-epreds-1}}

@@

In the present experiment, we investigated the extent to which
cognateness (phonological similarity between translation equivalents) is
sufficient for successful word translation, in the absence of semantic
knowledge. We tested two groups of monolingual British English natives
in a translation task that involved words in Catalan or Spanish, two
languages participants reported having no familiarity with. Participants
benefited strongly from cognateness, when the correct translation of the
presented words in Catalan or Spanish had few English phonological
neighbours with higher lexical frequency. This suggests that, in the
absence of distractors, even naïve participants can efficiently use
phonological similarity to succeed in a translation task. Our results
suggest that word-to-word connections at the phonological level might
play a strong role during L2 speech comprehension, specially in
low-proficiency listeners.

@@

@@

\subsubsection{Participants}\label{participants-1}

Participants in Spain were contacted via announcements in Faculties, and
were compensated 5€ or an Amazon voucher for the same value.
Participants gave informed consent before providing any data and the
study was conducted in accordance with ethical standards of the
Declaration of Helsinki and the protocol was approved by the Drug
Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar
(2020/9080/I). Data collection took place from June 08th, 2020 to June
28th, 2020. We collected data from 33 Spanish native participants living
in Spain (\emph{Mean} = 21.85 years, \emph{SD} = 3, Range = 18-33, 28, 5
female).

\subsubsection{Stimuli}\label{stimuli-1}

Same list of Catalan stimuli as in Experiment 1.

\subsubsection{Procedure}\label{procedure-1}

Same as in Experiment 1.

\subsubsection{Data analysis}\label{data-analysis-1}

Same approach as in Experiment 1.

@@

\textbf{?@tbl-results-1} shows a summary of participants' accuracy
across Experiments 1, 2, and 3. MCMC chains in the model showed strong
evidence of convergence (\(\hat{R}<1.01\)) (see Appendix 2 for more
detailed model diagnostics). Overall, participants responded less
accurately to words with more CLPNs than to words with fewer CLPNs,
regardless of the amount of phonological similarity between the
presented word and its translation. This is indicated by the size of the
regression coefficient of the two-way interaction between
\emph{Cognateness} and \emph{CLPN} (\(\beta\) = -0.409, 95\% CrI =
{[}-0.895, 0.111{]}, \emph{p}(ROPE) = 0.081). As anticipated,
participants' performance benefited from an increase in
\emph{Cognateness} (\(\beta\) = 9.274, 95\% CrI = {[}8.473, 10.269{]},
\emph{p}(ROPE) = 0), while the number of \emph{CLND} had the opposite
effect (\(\beta\) = -0.069, 95\% CrI = {[}-0.353, 0.192{]},
\emph{p}(ROPE) = 0.492). Figure~\ref{fig-epreds-1} illustrates the
posterior of the average predictions of the model for words with
different values of \emph{Cognateness} and \emph{CLPN}.

@@

\captionsetup{labelsep=none}\includegraphics{manuscript_files/figure-pdf/fig-epreds-2-1.pdf}

@@

\caption{\label{fig-epreds-2}}

@@

\subsubsection{Participants}\label{participants-2}

@@

\captionsetup{labelsep=none}\includegraphics{manuscript_files/figure-pdf/fig-epreds-3-1.pdf}

>>>>

\setlength{\LTpost}{0mm}

\begin{longtable}{l|rrrc}

@@

\toprule
\multicolumn{1}{l}{} &  & \multicolumn{2}{c}{Age} &  \\ 
\cmidrule(lr){3-4}
\multicolumn{1}{l}{} & N\textsuperscript{\textit{1}} & Mean ± SD  & Range & L2 \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 35 (8) & $21.80$ ± $2.08$ & $18$–$26$ & Russian (1) \\ 
cat-ENG & 36 (4) & $21.72$ ± $2.25$ & $18$–$25$ & French (1), German (1), Italian (1), Punjabi (1), Several (1) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & 33 (12) & $21.85$ ± $3.00$ & $18$–$33$ & French (9), German (1), Italian (2) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & 32 (1) & $21.72$ ± $2.59$ & $18$–$26$ & German (2), Japanese (1) \\ 
cat-ENG & 32 (0) & $22.31$ ± $2.39$ & $18$–$26$ & Cantonese (1), Irish (1) \\ 
\bottomrule

@@

\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}Number of included participants (number of excluded participants.)\\
\end{minipage}

@@

We arranged two lists of input words to be presented to participants in
the auditory modality: one in Catalan and one in Spanish. Words in the
Catalan list were 5.02 phonemes long on average (\emph{SD} = 1.49, Range
= 2-8), and the orthographic forms of their English translations (which
participants had to type) were 5.12 characters long on average
(\emph{SD} = 1.56, Range = 3-9. Words in the Spanish list were 5.52
phonemes long on average (\emph{SD} = 1.47, Range = 3-9), and the
orthographic form of their English translations (which participants had
to type), were 5.29 characters long on average (\emph{SD} = 1.77, Range
= 3-12.

Participants listened to one audio file in each trial, each containing a
single word presented in isolation. The audio files were the same ones
used in child experiments conducted in the Laboratori de Recerca en
Infància of Universitat Pompeu Fabra (Barcelona, Spain). These audio
files were recorded by a proficient Catalan-Spanish female bilingual
from the Metropolitan Area of Barcelona in a child-directed manner.
Catalan and Spanish words were recorded at 44,100 Hz in separate files
in the same session, and then de-noised using Audacity and normalised at
peak intensity using Praat (\citeproc{ref-broersma2021praat}{Broersma \&
Weenink, 2021}). The average duration of the Catalan audio files was
1.24 (\emph{SD} = 0.19, Range = 0.8-1.58). The average duration of the
Catalan audio files was 1.16 (\emph{SD} = 0.15, Range = 0.78-1.53).

For each translation pair, we defined three predictors of interest: the
lexical frequency of the correct translation (\emph{Frequency}), the
phonological similarity between the presented (Catalan or Spanish) word
and their correct English translation (\emph{Cognateness}), and the
presented word's number of cross-language phonological neighbours
(\emph{CLPN}).

We included \emph{Frequency} as a nuance predictor, under the hypothesis
that---keeping other predictors constant---participants would translate
higher-frequency words more accurately and faster than lower-frequency
words. Lexical frequencies of correct translations were extracted from
SUBTLEX-UK (\citeproc{ref-van2014subtlex}{Van Heuven et al., 2014}), and
transformed to Zipf scores. Words in the stimuli list without a lexical
frequency value were excluded from data analysis (2 in the Catalan list,
5 in the Spanish list).

We calculated \emph{Cognatenesss}, our main predictor of interest, by
computing the Levenshtein similarity between the X-SAMPA transcriptions
of each pair of translations using the \texttt{stringdist} R package
(\citeproc{ref-van2014stringdist}{van der Loo, 2014}). The Levenshtein
distance computes the edit distance between two character strings---in
this case, phoneme transcriptions---by counting the number of additions,
deletions, and substitutions necessary to make both strings identical
(\citeproc{ref-levenshtein1966binary}{Levenshtein et al., 1966}). We
divided this edit distance by the number of characters included in the
longest X-SAMPA transcription of the translation pair. This results in a
proportion score, in which values closer to zero correspond to lower
Levenshtein distances between phonological transcriptions (i.e., higher
similarity), and values closer to 1 correspond to higher Levenshtein
distances (i.e., lower similarity). This transformation also help
account for the fact that the Levenshtein distance tends to increase
with the length of the transcriptions. For interpretability, we
subtracted this proportion from one, so that values closer to one
correspond to higher similarity between phonological transcriptions, and
values closer to zero correspond to lower similarity between
phonological transcriptions. For example, the \emph{table}
(teɪbəl)-\emph{mesa} (mesa) translation pair had a 17\% similarity,
while the \emph{train} (\texttt{trEIn})-\emph{tren} (tɾEn) translation
pair had a 60\% similarity. Figure 1 summarises the lexical frequency,
phonological neighbourhood density and phonological overlap of the words
included in the Catalan and the Spanish lists.

For each Catalan and Spanish word, we calculated the number of
\emph{CLPN} by counting the number of English words with same or higher
lexical frequency, and whose phonological transcription (in X-SAMPA
format) different in up to one phoneme from that of the presented
Catalan or Spanish word. Lexical frequencies and phonological
transcriptions were extracted from the multilingual database CLEARPOND
(\citeproc{ref-marian2012clearpond}{Marian et al., 2012})\footnote{{[}Phonological
  transcriptions in CLEARPOND were generated from eSPEAK,
  \url{http://espeak.sourceforge.net/}{]}}.

@@

\caption{\label{tbl-stimuli}Stimuli properties.}

@@

\toprule
\multicolumn{1}{l}{} & \multicolumn{2}{c}{Frequency} & \multicolumn{2}{c}{Cognateness} & \multicolumn{2}{c}{CLPN} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
\multicolumn{1}{l}{} & Mean ± SD & Range & Mean ± SD & Range & Mean ± SD & Range \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $5.88$ ± $0.57$ & $4.43$–$7.24$ & $0.13$ ± $0.18$ & $0.00$–$0.75$ & $0.19$ ± $0.79$ & $0$–$5$ \\ 
cat-ENG & $5.89$ ± $0.59$ & $4.43$–$7.27$ & $0.16$ ± $0.18$ & $0.00$–$0.67$ & $0.76$ ± $2.53$ & $0$–$15$ \\ 
cat-SPA & $5.79$ ± $0.63$ & $4.48$–$7.70$ & $0.38$ ± $0.26$ & $0.00$–$1.00$ & $0.87$ ± $2.08$ & $0$–$12$ \\ 
\midrule 
\midrule 
\textbf{Mean} & $5.85$ & — & $0.225$ & — & $0.606$ & — \\ 
\bottomrule

@@

Each trial started with a yellow fixation point presented during one
second on the centre of the screen over a black background. After one
second, the audio started playing while the dot remained being displayed
until the audio ended. Upon the offset of the fixation point and audio,
participants were prompted to write their answer by a
\texttt{\textgreater{}} symbol. Typed letters were displayed in the
screen in real time to provide visual feed-back to participants.
Participants were allowed to correct their answer. Then, participants
pressed the \texttt{RETURN/ENTER} key to confirm their answer and start
and new trial.

@@

{\centering \includegraphics{_assets/img/design.png}

@@

\caption{Schematic representation of a trial in the experimental task}

@@

\paragraph{Data processing}\label{data-processing}

After data collection, participants answers were manually coded into the
following categories: \emph{Correct}, \emph{Typo}, \emph{Wrong},
\emph{False friend}, \emph{Other}. Responses were coded as
\emph{Correct} if the provided string of characters was identical to the
orthographic form of the correct translation. Trials were coded as
\emph{Typo} if the participant provided a string of characters only one
edit distance (addition, deletion, or substitution) apart from the
orthographic form of the correct translation (e.g., ``pengiun'' instead
of ``penguin''), as long as the response did not correspond to a
distinct English word. Responses were coded as \emph{False friend} if
the participant provided a phonologically similar incorrect translation.
Responses were coded as \emph{Other} (see Data analysis section for more
details). Both \emph{Correct} and \emph{Typo} responses were considered
as correct, while \emph{Wrong} and \emph{False friend} responses were
considered as incorrect. \emph{Other} responses were excluded from data
analysis.

Trials in which participants took longer than 10 seconds to respond were
also excluded. Participants contributed a total of 9152 valid trials
(5694 in Catalan, 3458 in Spanish). The task took approximately 15
minutes to be completed.

\paragraph{Modelling approach and statistical
inference}\label{modelling-approach-and-statistical-inference}

@@

We collected data for a total of 6,446 trials completed by 72 unique
participants. Of those trials, 2,988 were provided by 36 unique
participants who listened to Catalan words, and 3,458 trials were
provided by 36 unique participants who listened to Spanish words. We
excluded trials in which participants did not enter any text (\emph{n} =
72), in which a response in a language other than English was provided
(e.g., \texttt{agua}, \emph{n} = 51), in which participants did not
provide a whole word (e.g., \texttt{f}, \emph{n} = 5), and in which
participants added comments to the experimenter (e.g., \texttt{unsure},
(\emph{n} = 13). In addition, we excluded data from participants that
self-rated their oral and/or written skills in Catalan and Spanish, or
any other second language as four or higher in a five-point scale
(\emph{n} = 2), were diagnosed with a language (\emph{n} = 1), or did
not contribute more than 80\% of valid trials (\emph{n} = 9).

After applying trial-level and participant-level inclusion criteria, the
resulting dataset included 5,204 trials provided by 54 unique
participants. Of those trials, 2,602 were provided by 27 unique
participants who listened to Catalan words, and 2,604 trials were
provided by 32 unique participants who listened to Spanish words.
Responses given by English participants to Catalan presented words were
5.35 characters long on average (\emph{Median} = 5, \emph{SD} = 1.79,
Range = 1-14), while their translations to Spanish responses were 5.57
characters long on average (\emph{Median} = 5, \emph{SD} = 1.97, Range =
2-21).

@@

\begin{longtable}{l|rrrrrrrr}

@@

\toprule
\multicolumn{1}{l}{} & \multicolumn{4}{c}{Correct responses} & \multicolumn{4}{c}{Incorrect responses} \\ 
\cmidrule(lr){2-5} \cmidrule(lr){6-9}
\multicolumn{1}{l}{} & Correct & (\%) & Typo & (\%) & Wrong & (\%) & False friend & (\%) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{9}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
cat-ENG & $429$ & ($16.47$) & $11$ & ($0.42$) & $2,082$ & ($79.95$) & $82$ & ($3.15$) \\ 
spa-ENG & $374$ & ($14.37$) & $2$ & ($0.08$) & $2,117$ & ($81.36$) & $109$ & ($4.19$) \\ 
\midrule 
\emph{Sum} & $803$ & — & $13$ & — & $4,199$ & — & $191$ & — \\ 
\emph{Mean} & — & $15.42$ & — & $0.25$ & — & $80.66$ & — & $3.67$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{9}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & $780$ & ($46.93$) & $20$ & ($1.20$) & $736$ & ($44.28$) & $126$ & ($7.58$) \\ 
\midrule 
\emph{Sum} & $780$ & — & $20$ & — & $736$ & — & $126$ & — \\ 
\emph{Mean} & — & $46.93$ & — & $1.20$ & — & $44.28$ & — & $7.58$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{9}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
cat-ENG & $477$ & ($18.01$) & $7$ & ($0.26$) & $1,986$ & ($75.00$) & $178$ & ($6.72$) \\ 
spa-ENG & $590$ & ($19.53$) & $6$ & ($0.20$) & $2,294$ & ($75.94$) & $131$ & ($4.34$) \\ 
\midrule 
\emph{Sum} & $1,067$ & — & $13$ & — & $4,280$ & — & $309$ & — \\ 
\emph{Mean} & — & $18.77$ & — & $0.23$ & — & $75.47$ & — & $5.53$ \\ 
\bottomrule

@@

\begin{longtable}{l|rrrrrrrrr}

@@

\toprule
\multicolumn{1}{l}{} &  & \multicolumn{4}{c}{Accuracy (\%)} & \multicolumn{4}{c}{Valid trials} \\ 
\cmidrule(lr){3-6} \cmidrule(lr){7-10}
\multicolumn{1}{l}{} & N & Mean & SD & SE & Range & Mean & N trials & SD & Range \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{10}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $27$ & $15.86$ & $5.20$ & $3.05$ & $8.82$–$28.71$ & $96.37$ & $2,602$ & $2.88$ & $87$–$98$ \\ 
cat-ENG & $32$ & $18.48$ & $4.89$ & $3.27$ & $10.84$–$32.56$ & $81.38$ & $2,604$ & $3.17$ & $71$–$83$ \\ 
\midrule 
\emph{Sum} & $59$ & — & — & — & — & — & $5,206$ & — & $158$ \\ 
\emph{Mean} & — & $17.17$ & $5.05$ & $3.16$ & — & $8,887.27$ & — & $302.71$ & — \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{10}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & $21$ & $48.30$ & $5.29$ & $10.54$ & $38.27$–$58.97$ & $79.14$ & $1,662$ & $3.14$ & $72$–$82$ \\ 
\midrule 
\emph{Sum} & $21$ & — & — & — & — & — & $1,662$ & — & $72$ \\ 
\emph{Mean} & — & $48.30$ & $5.29$ & $10.54$ & — & $7,914.29$ & — & $313.51$ & — \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{10}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $31$ & $20.92$ & $8.29$ & $3.76$ & $5.88$–$44.12$ & $97.45$ & $3,021$ & $1.80$ & $88$–$98$ \\ 
cat-ENG & $32$ & $19.74$ & $4.94$ & $3.49$ & $10.34$–$27.91$ & $82.75$ & $2,648$ & $0.44$ & $82$–$83$ \\ 
\midrule 
\emph{Sum} & $63$ & — & — & — & — & — & $5,669$ & — & $170$ \\ 
\emph{Mean} & — & $20.33$ & $6.62$ & $3.62$ & — & $9,010.08$ & — & $112.22$ & — \\ 
\bottomrule

@@

\includegraphics{manuscript_files/figure-pdf/fig-epreds-1-1.pdf}

@@

\caption{\label{fig-epreds-1}Posterior model-predicted mean accuracy in
Experiment 1. Predictions were generated from 4,000 posterior samples,
extracted for different values of \emph{CLPN} (0, 2, 4, 8, 12) and
\emph{Cognateness} (1-100). Predictions are plotted separately for
English participants translating Catalan words, and for English
participants translating Spanish words. Lines indicate mean predictions,
and intervals indicate 95\%, 89\%, 78\%, 67\%, and 50\% credible
intervals (CrI).}

@@

In the present experiment, we investigated the extent to which
cognateness (phonological similarity between translation equivalents) is
sufficient for successful word translation, in the absence of semantic
knowledge. We tested two groups of monolingual British English natives
in a translation task that involved words in Catalan or Spanish, two
languages participants reported having no prior familiarity with.
Participants benefited strongly from cognateness, when the correct
translation of the presented words in Catalan or Spanish had few English
phonological neighbours with higher lexical frequency. This suggests
that, in the absence of distractors, even naïve participants can
efficiently use phonological similarity to succeed in a translation
task. Our results suggest that word-to-word connections at the
phonological level might play a strong role during L2 speech
comprehension, specially in low-proficiency listeners.

@@

Results in Experiment 1 suggest that English natives were able to
exploit the phonological similarity between unfamiliar words in Catalan
and Spanish to provide accurate translations to English. English, a
Germanic language, is relatively distant from Catalan and Spanish, two
Romance languages. English shares fewer cognates with Catalan and
Spanish than it does with typologically closer languages, like Dutch and
German. In Experiment 2, we investigated whether listeners of an
unfamiliar but typologically closer language benefit more strongly from
phonological similarity when performing the same task as in Experiment
1. To this aim, we presented Spanish participants, who reported
little-to-no prior familiarity with Catalan, with Catalan words.

@@

Participants in Spain were contacted via announcements in Faculties, and
were compensated 5€ or an Amazon voucher for the same value.
Participants gave informed consent before providing any data and the
study was conducted in accordance with ethical standards of the
Declaration of Helsinki and the protocol was approved by the Drug
Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar
(2020/9080/I). Data collection took place from June 08th, 2020 to June
28th, 2020. We collected data from 33 Spanish native participants living
in Spain (\emph{Mean} = 21.85 years, \emph{SD} = 3, Range = 18-33, 28, 5
female). Stimuli were the same list of Catalan stimuli as in Experiment
1. Procedure and data analysis were identical as in Experiment 1.

@@

MCMC chains in the model showed strong evidence of convergence
(\(\hat{R}<1.01\)) (see Appendix 2 for more detailed model diagnostics).
Overall, participants responded less accurately to words with more CLPNs
than to words with fewer CLPNs, regardless of the amount of phonological
similarity between the presented word and its translation. This is
indicated by the size of the regression coefficient of the two-way
interaction between \emph{Cognateness} and \emph{CLPN} (\(\beta\) =
-0.409, 95\% CrI = {[}-0.895, 0.111{]}, \emph{p}(ROPE) = 0.081). As
anticipated, participants' performance benefited from an increase in
\emph{Cognateness} (\(\beta\) = 9.274, 95\% CrI = {[}8.473, 10.269{]},
\emph{p}(ROPE) = 0), while the number of \emph{CLND} had the opposite
effect (\(\beta\) = -0.069, 95\% CrI = {[}-0.353, 0.192{]},
\emph{p}(ROPE) = 0.492). Figure~\ref{fig-epreds-1} illustrates the
posterior of the average predictions of the model for words with
different values of \emph{Cognateness} and \emph{CLPN}.

@@

\includegraphics{manuscript_files/figure-pdf/fig-epreds-2-1.pdf}

@@

\caption{\label{fig-epreds-2}Posterior model-predicted mean accuracy in
Experiment 1. Predictions were generated from 4,000 posterior samples,
extracted for different values of \emph{CLPN} (0, 2, 4, 8, 12) and
\emph{Cognateness} (1-100). Predictions are plotted separately for
English participants translating Catalan words, and for English
participants translating Spanish words. Lines indicate mean predictions,
and intervals indicate 95\%, 89\%, 78\%, 67\%, and 50\% credible
intervals (CrI).}

@@

@@

\includegraphics{manuscript_files/figure-pdf/fig-epreds-3-1.pdf}
